\sick, the spectroscopic inference crank

### 2014-08-09 document/manuscript.tex
Updated manuscript

% TODO:
% [ ] Create example code figures for the model
%  -- use pygmentize when necessary
% [X] Create some awesome GC spectral plots with Seaborn
%  -- [X] uncertainty in parameters as a function of S/N
%  -- [ ] Box plot distribution for the clusters
% [X] Run Sol at high-resolution?
%  -- [ ] plot the Sun
%  -- [ ] plot of random samples + optimization
%  -- [ ] plot triangle contours for the Sun

\begin{abstract}
In this \article{} I introduce \sick{}, the spectroscopic inference crank, an open source probabilistic code for inferring astrophysical parameters from spectra. \sick{} enables any user to easily construct an approximate \textit{generative model} for spectral data, allowing for precise inference of quantifiable properties for a host of astrophysical processes. Model fluxes (or intensities) are approximated by efficient multi-dimensional interpolation of pre-computed spectra. This allows any user to capitalise on the plethora of published synthetic and observed spectra available without extra effort. Additional phenomena that transform the data (e.g., redshift, continuum, smoothing) are incorporated as free parameters. Outlier pixels (e.g., cosmic rays or poorly modelled regimes) are treated with a mixture model, and a simplistic noise model is included to account for systematically underestimated variance. Combining these phenomena into a justified, quantitative model allows for precise inference with credible uncertainties. All aspects of the probabilistic model are introduced, and the implementation details are described. I demonstrate the utility of a generative model approach by precise stellar photospheric measurements from noisy (S/N $\sim{} 7$) high- and low-resolution spectra. \sick{} is very easy to use, well-tested, parallelised, and freely available online through GitHub under the MIT license. 
\end{abstract}

\section{Introduction}
Most of what we know about astrophysics has been interpreted from spectra. Given how informative spectroscopic data is to our understanding of astrophysics, it is not surprising that the last decade has seen a substantial increase of publicly accessible spectra. Large scale surveys have helped drive this trend, each releasing in excess of hundreds of thousands \citep[e.g.,][]{wigglez,boss,segue,rave} or even millions \citep{lamost,galah} of spectra.

The spectra are chosen using different source selections, and the data vary in wavelength coverage, resolution, and noise distributions. For these reasons many collaborations expend significant resources to produce bespoke analysis software to achieve their scientific objectives. Unfortunately this impedes scientific reproducibility, as many codes still remain closed-source more than a decade after the original article was published. Subsequently any comprehensive literature comparison effectively becomes impossible, as systematics can be difficult to accurately characterise without in-depth knowledge of the method or access to the software.

Broadly speaking, there are three types of methods employed for spectral analysis: measuring strength of spectral features, data-generating models or template-matching methods. Approaches that measure spectral features are inexpensive, but regularly encounter problems with blended (often hidden) lines or continuum placement. As such, some subjective interaction or tuning is often required. Data-generative methods compute model spectra at run-time, and whilst accurate, can be prohibitively expensive for large samples. For template-matching methods, synthetic spectra are only generated once to produce a grid of fluxes or intensities for each permutation of astrophysical quantities. Although there are differences between these methods, the preparatory steps are usually the same. Spectra are placed at rest-frame by calculating line-of-sight velocities, typically by cross-correlation, before being continuum-normalised to flux intensities.

Credible uncertainties can be difficult to discern from these approaches. The implied assumptions of a $\chi^2$ distribution are often broken, and uncertainties in the doppler-shift, smoothing, and continuum normalisation steps are almost always ignored. These effects result in ill-characterised uncertainties in astrophysical parameters. Alternatively the uncertainties are frequently, and often incorrectly, assumed to be dominated by systematic effects, and the magnitude of this effect is gauged by comparisons with a `trustworthy' literature sample. The simplest procedure is to then state all objects have roughly the same uncertainties in astrophysical parameters, since they are assumed to be dominated by systematics. This is an incorrect approach: there are few, if any, examples of homoscedastic datasets in astrophysics. The noise properties of each spectrum \textit{are} different, and the parameter uncertainties (random and systematic) will differ for every object. Consequently the uncertainties in astrophysical parameters by template-matching methods are generally found to either be incorrectly assumed, underestimated, or at least ill-characterised.

In addition to affecting the uncertainties, the effects of redshift, continuum normalisation and smoothing \textit{will} systematically bias the reported maximum-likelihood parameters. For example, there are a number of controversial examples within the literature where the subjective (human) decision of continuum placement has significantly altered the scientific conclusions \citep[e.g., see][where this issue is argued in great detail]{kerzendorf}. The implications of these phenomena \textit{must} be considered if we are to understand subtle astrophysical processes. Spectroscopy requires objectivity: one should endeavour to incorporate these phenomena as free parameters into a generative model and infer them simultaneously with our astrophysical parameters.

In this \article{} I present \sick{}: a well-tested, MIT-licensed probabilistic software package for inference from spectroscopic data. \sick{} employs an approximation to data-generating models: instead of attempting to model all manners of expensive astrophysical processes (e.g., supernova explosions, stars, and any other astrophysically interesting process) at runtime, it performs efficient multi-dimensional interpolation of model spectra from pre-computed grids. This approach is suitable for a plethora of different astrophysical processes, allowing any user to easily specify a model using any existing published spectral grid, and infer astrophysical properties from their data. The probabilistic model is described in Section \ref{sec:model}. A number of examples are included in Section \ref{sec:examples}, where I present accurate and precise objective photospheric measurements using noisy spectra. I conclude in Section \ref{sec:conclusions} with references to the online documentation and future applicability of the software.

\section{The Generative Model}

The primary advantage of \sick{} is the simultaneous inference of both astrophysical and other pertinent parameters that can alter the spectra. The generative model described here is agnostic as to \textit{what} the astrophysical parameters actually describe. Typical examples might be properties of supernova (e.g., explosion energies and luminosities), galaxy characteristics from integrated light, or mean plasma properties of a stellar photosphere. There are a plethora of synthetic spectral grids published for these types of applications \citep{who;who;who}. All of these models are fully-\sick{} compatible. However given the research background of the author, I will introduce the probabilistic model by considering the problem of inferring mean photospheric properties from a stellar spectrum.

The first step of the (approximate) generative model is to interpolate a model spectrum. \sick{} employs the Qhull algorithm \citep{qhull} to linearly interpolate in multiple dimensions. Irregularly-spaced grids are perfectly acceptable; Qhull does not require the grid points to be rectangular or regularly spaced. Large grids of model spectra can be cached for computational efficiency (and this is recommended), allowing for the total size of the model grid to far exceed the available random access memory.\footnote{In practice this is performed through memory-mapping.} I introduce an interpolation function $S(\bm{\psi})$ that produces a model spectrum with $J$ points $\{\lambda_j,I_j\}_{0}^{J}$ at any evaluable point $\bm{\psi} \equiv (T_{\rm eff},\log{g},{\rm [M/H]},[\alpha/{\rm Fe}])$ within the grid boundaries.

@@

The doppler shift of the source is modelled by transforming the model spectrum (in uniformly sampled $\log\lambda$ space). For computational efficiency \sick{} solves for redshift $z$, and returns posterior distributions in the user's preferred units (e.g., km s$^{-1}$).

\begin{equation}
\lambda_{s} = \lambda_{j}(1 + z)
\end{equation}

After the model spectrum is smoothed and shifted, model intensities are required at the wavelengths $\lambda_i$ of the observed pixels. The complete function to yield model intensities $M_i$ at wavelengths $\lambda_i$ is therefore a function of the parameters of interest $\bm{\psi}$, redshift $z$, smoothing kernel $\sigma_{s}$, and the wavelengths of the observed pixels $\lambda_i$:

@@

In principle there is no reason to believe that the doppler-shifted, smoothed model intensities $M_i$ should match the data at all. The counts and shape of observed fluxes are a function of source magnitude, exposure time, instrument sensitivities, atmospheric conditions, and a host of unaddressed effects. In contrast the model spectra are calculated either as intensities (e.g., $I_j$) or calibrated fluxes to an empirical system. Even for the true values of $\bm{\psi}$, a function is required to normalise\footnote{The normalisation process is frequently abused by stellar spectroscopists in the literature. Wherever possible, data should not be transformed. One should seek to fit a model to the data, not the other way around.} the model to the data. As such I transform the \textit{model} by some function $C$ to fit the data. Although the function $C$ incorporates a number of effects (e.g., source blackbody temperature, dust, instrument sensitivities), they are phenomena that typically cannot be separated without additional information, and here I only care about their combined effect. I wish to ensure that the overall shape of the data is accounted for, which is usually achievable with a low-order polynomial

@@

\noindent{}with $m$ free coefficients $c_{0\dots{}m}$, where $m$ is specified by the user. Other continuum functions are available in \sick{}, and the user is encouraged to experiment. This allows us to express the \textit{expected flux} $E_i$ at a given observed pixel with wavelength $\lambda_i$ as:

@@

In many astrophysical cases the flux uncertainties $\sigma_{i}$ for a given pixel are not well-characterised. This is often due to unpropagated uncertainties during data reduction. Since the observed flux counts are usually Poisson distributed, in many astrophysical scenarios the pixel flux uncertainty can be made estimated as $\sigma_i \sim 1/\sqrt{F_{i}}$. Knowing that this approximation may under-estimate the noise, I include an additional parameter to account for the possibility that the variance in all pixels is underestimated by some fractional amount $f$ (Eq. \ref{eq:noise-model}). While this crude noise model is probably unrepresentative of the data in a large number of cases, including \textit{any} noise model is preferable to none.

\begin{equation}
\sigma_{t_i}^2 = \sigma_{i}^2(1 + f)
\label{eq:noise-model}
\end{equation}

@@

\begin{equation}
p\left(F_i|\lambda_i,\sigma_{i},\bm{\psi},z,\sigma_{s},\{c\}_{0}^{m},f\right) = 
 \frac{1}{\sqrt{2\pi\sigma_{t_i}^2}}\exp{\left(-\frac{\left[F_i - E_i\right]^2}{2\sigma_{t_i}^2}\right)}
 \label{eq:p_model}
\end{equation}

@@

From the original astrophysical parameters $\bm{\psi}$ that we care about, we now have an additional $3 + m$ parameters to consider. The situation seemingly becomes more complex when we consider separately observed channels (also colloquially expressed as beams or apertures). Many spectrographs provide small portions of spectra (a channel) separated by some large gap. The data for each channel are generally captured by different CCDs and are therefore reduced separately.

@@

Finally, I consider the handling of outliers in the data. These may be in the form of cosmic ray spikes, improper calibration of the data (e.g., skylines) or poorly modelled spectral regions. Treatment of these artefacts is achieved with a Gaussian mixture model, a combination of two models. This requires the inclusion of two (final) additional parameters: $P_o$ and $V_o$. In the mixture model, the data are fit by the sum of amplitudes ($1 - P_o$ and $P_o$, respectively) of two distributions: the expected fluxes $E_i$, and a normal distribution centered along the continuum function $C_i$ with variance $\sigma_{t_i}^2 + V_{o}$. The prior distribution function $p\left(V_o\right)$ requires $V_{o}$ to always be positive (Eq. \ref{eq:default_priors}), and as such the outlier distribution will \textit{always} have a larger variance. Because distributions of smaller variance are more informative, a fit to the expected fluxes $E_{i}$ is generally preferred wherever possible. For brevity I define $\bm{\kappa} \equiv (\bm{\psi},\{z,\sigma_s,\{c_k\}_{k=0}^{m},f\}_{0}^{N_{c}})$, and the likelihood for the mixture model is given by

@@

\noindent{}where $p_{model}$ refers to $p$ in Eq. \ref{eq:p_model} and

\begin{equation}
p_{outlier}\left(F_i|\lambda_i,\sigma_i,\bm{\kappa},V_{o},P_o\right) = \frac{1}{\sqrt{2\pi\left(\sigma_{t_i}^2 + V_{o}^2\right)}} \exp\left(-\frac{[F_i - C_i]^2}{2\left[\sigma_{t_i}^2 + V_{o}^2\right]}\right)
\end{equation}

@@

\begin{equation}
\mathcal{L} = \prod_{i=1}^{N} \left[ \frac{1-P_o}{\sqrt{2\pi\sigma_{t_i}^2}}\,\exp\,\left(-\frac{[F_i - E_i]^2}{2\sigma_{t_i}^{2}}\right) + \frac{P_o}{\sqrt{2\pi\left[\sigma_{t_i}^2 + V_o\right]}}\,\exp\,\left(-\frac{[F_i - C_i]^2}{2\left[\sigma_{t_i}^{2} + V_o\right]}\right)\right]
\label{eq:full_likelihood}
\end{equation}

I define the full parameter space with $\bm{\theta} \equiv \left(\bm{\psi},\{z,\sigma_s,\{c_{b_k}\}_{k=0}^{m},f\}_{b=0}^{N_{c}},V_o,P_o\right)$. From Bayes theorem we obtain the posterior probability distribution for $\bm{\theta}$ (up to a constant) as
\begin{eqnarray}
\mathcal{P} & \propto & likelihood \times prior \nonumber \\
p(\bm{\theta}|\{F_i\}_{i=1}^{N}) & \propto & p(\{F_i\}_{i=1}^{N}|\bm{\theta})\,\times\,p(\bm{\theta})
\label{eq:probability}
\end{eqnarray}

\noindent{}where $p(\bm{\theta}|\{F_i\}_{i=1}^{N})$ is the probability $\mathcal{P}$ of $\bm{\theta}$ given the data (and the model: see Section \ref{sec:sun}), $p(\{F_i\}_{i=1}^{N}|\bm{\theta})$ is our previously defined likelihood function $\mathcal{L}$ (Eq. \ref{eq:full_likelihood}), and $p(\bm{\theta})$ is the prior probability distribution. Priors are discussed in more detail in Section \ref{sec:priors}.

%

@@

With all of the combined effects, there are 16 free parameters including $\bm{\psi}$, our astrophysical quantities of interest. For the examples presented in Section \ref{sec:examples}, this requires interpolation between $\sim{}92\times10^9$ pixels in four dimensions. While the description might appear daunting, the problem is tractable, numerically efficient, and easy to configure. By default \sick{} numerically solves this problem in three sequential steps, which are described in the following sections: scattering, optimisation, and Monte-Carlo Markov Chain (MCMC) sampling.

@@

We seek to maximise $\mathcal{P}$ (or in practice, $\log\mathcal{P}$) and calculate the posterior probability distributions for $\bm{\theta}$ given the data. In the first step we calculate the log-probability $\log{(\mathcal{P})}$ for $N_{sample}$ randomly drawn points from all over the astrophysical parameter space $\bm{\psi}$. Since \sick{} allows for arbitrarily large parameter spaces in $N_{D}$ dimensions, initially sampling the parameter space $N_{sample}$ times provides a coarse representation of probability. If the probability distribution function were smoothly distributed across all $\bm{\psi}$, or the optimisation step was sufficiently robust for all potential applications, then in principle \textit{any} single point would be an adequate starting guess. However there is no way of knowing \textit{a priori} how smooth the probability landscape is for a given problem.

The $N_{samples}$ points are uniformly drawn in $\bm{\psi}$ by default, unless priors are specified (see Section \ref{sec:priors}). A model spectrum $S(\bm{\psi})$ is interpolated for each point, and is used to estimate the redshift and continuum parameters. After smoothing the model fluxes $I_j$ by $\sigma_s$ (as drawn from an explicit prior, or estimated by $\left|\mathcal{N}\left(0, 1\right)\right|$) the smoothed spectrum is cross-correlated with the data to yield a redshift $z$. Continuum parameters $\{c_k\}_{k=0}^{m}$ are similarly estimated by fitting a function (e.g., a polynomial) to the data divided by the smoothed flux at each $\lambda_i$. Estimating continuum and redshift provides us with a reasonable guess of the probability for any randomly scattered point $\bm{\psi_i}$. If outlier modelling is included in the model then $P_o$ is distributed as $\mathcal{U}\left(0, 1\right)$ by default, and $V_o$ is estimated by $\mathcal{N}\left(\widetilde{F_i}, \frac{1}{2}\widetilde{F_i}\right)$ and $\mathcal{N}\left(\mbox{Var}\left(F_i\right),\frac{1}{2}\mbox{Var}\left(F_i\right)\right)$ respectively. This procedure is \textit{only} employed for the initial random scattering stage: it is \textit{not} applicable for MCMC sampling.

@@

After the random scattering step, the most probable $\bm{\theta}$ point is used as an initial guess for numerical optimisation. A number of suitable minimisation algorithms are available in \sick{} through the SciPy optimization \cite{scipy,scipy-optimize} module. The Nelder-Mead algorithm \citep{nelder-mead} is employed by default, and the full suite of optimisation algorithms available is listed in the online documentation. Unlike other optimisation techniques, the Nelder-Mead algorithm does not approximate first- or second-order derivatives, and can consequently be less efficient than other approaches. However it is robust in high-dimensional space, even in the presence of substantial noise, and has been successfully employed in a wide range of engineering and scientific problems. The reader is encouraged to experiment with other optimisation algorithms if the Nelder-Mead algorithm proves unsuitable or takes an untenable amount of time. Since these optimisation algorithms are minimisation techniques and we seek to maximise the log-probability $\log{\left(\mathcal{P}\right)}$, we numerically optimise the parameters $\bm{\theta}$ by minimising the negative log-probability $-\log{\left(\mathcal{P}\right)}$.

@@

The random scattering and numerical optimisation steps efficiently provide an accurate estimate of the optimal parameters $\bm{\theta}$. Once the optimisation step is complete \sick{} employs the affine-invariant Metropolis-Hastings MCMC ensemble sampler proposed by \citet{goodman;weare}, and implemented by \citet{emcee}. If the optimisation steps is not employed, initial points for the MCMC walkers are drawn in the same way as described in the scattering process. When the optimisation step is used, the initial points are taken from a small multi-dimensional ball around the optimised parameters $\bm{\theta_{opt}}$.

% [TODO] "Experiment" with proposal widths.

Priors represent our initial knowledge about a particular parameter before looking at the data, and are necessary for any Bayesian analysis. A number of different prior distributions can be specified by the user in the \sick{} model configuration file. When no prior is specified, the following uninformative prior distributions are assumed (for all channels, where appropriate):

@@

A consequence of allowing irregular model grids is that a model spectrum cannot be interpolated for some values of $\bm{\psi}$, even if they fall within $\left(\min\left[\bm{\psi}\right], \max\left[\bm{\psi}\right]\right)$. In these cases $p\left(\bm{\psi}\right) = 0$ and thus $\log\left(\mathcal{P}\right) = -\infty$.

>>>>

\begin{abstract}
In this \article{}  I  introduce \sick{}, the spectroscopic inference crank, an open source probabilistic code for inferring astrophysical parameters from spectra. \sick{} enables any user to easily construct an approximate \textit{generative model} for spectral data, allowing for precise inference for a host of astrophysical quantities. Model fluxes (or intensities) are approximated by efficient multi-dimensional interpolation of pre-computed spectra. This allows any user to capitalise on the plethora of published synthetic and observed spectral libraries with minimal effort. Additional phenomena that transform the data (e.g., redshift, continuum, smoothing) are incorporated as free parameters. Outlier pixels (e.g., cosmic rays or poorly modelled regimes) can be treated with a mixture model, and a simplistic noise model is included to account for systematically underestimated variance. Combining these phenomena into a scalar-justified, quantitative model allows for precise inference with credible uncertainties. Commonly employed features are introduced, and the implementation details are described. I demonstrate the utility of a generative model approach with accurate and precise stellar photospheric measurements from noisy (e.g., S/N $\sim{} 7$) high- and low-resolution spectra. \sick{} is very easy to use, well-tested, parallelised, and freely available online through GitHub under the MIT license. 
\end{abstract}

\section{Introduction}
Most of our understanding of astrophysics has been interpreted from spectra. Given how informative spectroscopic data is to our understanding of astrophysics, it is not surprising that the last decade has seen a substantial increase of publicly accessible spectral data. Large scale surveys have driven this trend, each releasing in excess of hundreds of thousands \citep[e.g.,][]{wigglez,boss,segue,rave,gaia-eso} of spectra. Millions more spectra are expected in the coming years \citep[e.g.,][]{lamost,galah}.

The spectra are obtained from different astrophysical sources to meet specific scientific objectives, and the data vary in wavelength coverage, resolution, and noise distributions. For these reasons many collaborations expend significant resources to produce bespoke analysis software. Unfortunately this impedes scientific reproducibility, as many codes still remain closed-source more than a decade after the original article was published. Any comprehensive literature comparison subsequently becomes impossible, as systematics can be difficult to accurately characterise without in-depth knowledge of the methods or access to the software.

Broadly speaking there are three types of methods employed for spectral analysis: measuring the strengths of spectral features, pure data-generating models or template-matching methods. Approaches that measure spectral features are inexpensive, but regularly encounter problems with blended (often hidden) lines or continuum placement. As such, some subjective interaction, tuning, or ad-hoc `calibration' is almost always required. Data-generating methods compute model spectra at run-time, and whilst accurate, can be prohibitively expensive for large samples. For template-matching methods, synthetic spectra are only generated once to produce a grid of fluxes or intensities for a subset of permutations of astrophysical quantities. Although there are differences between these methods, the preparatory steps are usually the same. Spectra are placed at rest-frame by calculating line-of-sight velocities, typically by cross-correlation, before being continuum-normalised to flux intensities.

Credible uncertainties can be difficult to discern from these approaches. The implied assumptions of a $\chi^2$ distribution are often inapplicable, and uncertainties in the doppler-shift, smoothing, and normalisation steps are almost always ignored. These effects result in ill-characterised uncertainties in astrophysical parameters. Alternatively the uncertainties are frequently assumed to be approximately the same for all objects.  This is an incorrect approach: there are few, if any, examples of homoscedastic datasets in astrophysics. The noise properties of each spectrum \textit{are} different, and the parameter uncertainties (random and systematic) will differ for every object. Consequently the uncertainties in astrophysical parameters by template-matching methods are generally found to either be incorrectly assumed, under-estimated, or at least ill-characterised.

In addition to affecting the uncertainties, the effects of redshift, continuum normalisation and smoothing \textit{will} systematically bias the reported maximum-likelihood parameters. Experienced spectroscopists will frequently differ in their decision of continuum placement, even in the most straightforward cases (e.g., metal-poor stars). For example, there are a number of controversial examples within the literature where the subjective (human) decision of continuum placement have significantly altered the scientific conclusions \citep[e.g., see][where this issue is discussed in great detail]{kerzendorf}. The implications of these phenomena \textit{must} be considered if we are to understand subtle astrophysical processes. Spectroscopy requires objectivity: one should endeavour to incorporate these phenomena as free parameters into a generative model and infer them simultaneously with the astrophysical parameters.

In this \article{} I present \sick{}: a well-tested, MIT-licensed probabilistic software package for inference from spectroscopic data. \sick{} employs an approximation to data-generating models. Instead of attempting to model all kinds of expensive astrophysical processes (e.g., supernova explosions, stars, and any other interesting astrophysical processes) at run-time, it performs efficient multi-dimensional interpolation of model spectra from pre-computed grids. This approach is suitable for a plethora of different astrophysical processes, allowing any user to easily specify a model using existing published spectral grids, and infer astrophysical properties from their data. Aspects of the probabilistic model are described in Section \ref{sec:model}. A number of examples are included in Sections \ref{sec:inference-test} and \ref{sec:examples}, where I present accurate and precise objective photospheric measurements using noisy spectra. I conclude in Section \ref{sec:conclusions} with references to the online documentation and applicability of the software.

\section{The Generative Model}
\label{sec:model}

The primary advantage of \sick{} is the simultaneous inference of both astrophysical and other pertinent parameters that can alter the spectra. The generative model described here is agnostic as to \textit{what} the astrophysical parameters actually describe. Typical examples might be properties of supernova (e.g., explosion energies and luminosities), galaxy characteristics from integrated light, or mean plasma properties of a stellar photosphere. There are a plethora of spectral libraries (observed and synthetic) published for these types of applications \citep[e.g.,][]{snid,pegase,phoenix,pollux}. All of these models are fully-\sick{} compatible. However given the research background of the author, I will introduce the probabilistic model by focussing on stellar photospheric inferences.

The first step of the (approximate) generative model is to interpolate a model spectrum. \sick{} employs the Quickhull algorithm \citep{quickhull} to linearly interpolate in multiple dimensions. Irregularly-spaced grids are perfectly acceptable; Quickhull does not require the grid to be rectangular or regularly-spaced points. Large grids of model spectra are cached for computational efficiency, allowing for the total size of the model grid to far exceed the available random access memory.\footnote{In practice this is performed by efficient memory-mapping.} I introduce an interpolation function $S(\bm{\psi})$ that produces a model spectrum with $J$ points $\{\lambda_j,I_j\}_{0}^{J}$ at any evaluable point $\bm{\psi} \equiv (T_{\rm eff},\log{g},{\rm [M/H]},[\alpha/{\rm Fe}])$ within the grid boundaries.

@@

The doppler shift of the source is treated by transforming the model spectrum (in uniformly sampled $\log\lambda$ space). \sick{} solves for redshift $z$ (Equation \ref{eq:redshift}), and returns posterior distributions in the user's preferred units (e.g., km s$^{-1}$).

\begin{equation}
\label{eq:redshift}
\lambda_{s} = \lambda_{j}(1 + z)
\end{equation}

After the model spectrum is smoothed and doppler shifted, model intensities are required at the wavelengths $\lambda_i$ of the observed pixels. The complete function to yield model intensities $M_i$ at wavelengths $\lambda_i$ is therefore a function of the parameters of interest $\bm{\psi}$, redshift $z$, smoothing kernel $\sigma_{s}$, and the wavelengths of the observed pixels $\lambda_i$:

@@

In principle there is no reason to believe that the doppler-shifted, smoothed model intensities $M_i$ should match the data at all. The counts and shape of observed fluxes are a function of source magnitude, exposure time, instrument sensitivities, atmospheric conditions, and a host of unaddressed effects. In contrast the model spectra are calculated either as intensities (e.g., $I_j$) or fluxes calibrated to an empirical system. Even for the true values of $\bm{\psi}$, a function is required to normalise\footnote{The normalisation process is frequently abused by stellar spectroscopists in the literature. Wherever possible, data should not be transformed. One should seek to fit a model to the data, not the other way around.} the model to the data. As such I transform the \textit{model} intensities by some function $C$ to fit the data. Although the function $C$ incorporates a number of effects (e.g., source blackbody temperature, dust, instrument sensitivities), they are phenomena that typically cannot be separated without additional information, and here I only care about their combined effect. I only wish to ensure that the overall shape of the data are accounted for, which is usually achievable with a low-order polynomial

@@

\noindent{}with $m$ free coefficients $c_{0\dots{}m}$, where $m$ is specified by the user. Other continuum functions are available in \sick{}, and the user is encouraged to experiment for a specific problem. This allows us to express the \textit{expected flux} $E_i$ at a given observed pixel with wavelength $\lambda_i$ as:

@@

In many astrophysical cases the flux uncertainties $\sigma_{i}$ for a given pixel are not well-characterised. This is often due to unpropagated uncertainties during data reduction. Since the observed flux counts are expected to be Poisson-distributed, in many astrophysical scenarios the pixel flux uncertainty can be made estimated as $\sigma_i \sim 1/\sqrt{F_{i}}$. Knowing that this approximation may underestimate the noise, I include an additional parameter to account for the possibility that the variance in all pixels is underestimated by some fractional amount $f$ (Equation \ref{eq:noise-model}). While this crude noise model is probably unrepresentative of the data in a large number of cases, including \textit{any} noise model is preferable to none.

\begin{equation}
s_{i}^2 = \sigma_{i}^2 + f^{2}E_{i}^2
\label{eq:noise-model}
\end{equation}

@@

\begin{equation}
p\left(F_i|\lambda_i,\sigma_{i},\bm{\psi},z,\sigma_{s},\{c\}_{0}^{m},f\right) = 
 \frac{1}{\sqrt{2\pi{}s_{i}^2}}\exp{\left(-\frac{\left[F_i - E_i\right]^2}{2s_{i}^2}\right)}
 \label{eq:p_model}
\end{equation}

@@

From the original astrophysical parameters $\bm{\psi}$ that I care about, I now have an additional $3 + m$ parameters to consider. The situation seemingly becomes more complex when separately observed channels are considered. Many spectrographs provide small portions of spectra (a channel, also colloquially expressed as beams or apertures) separated by some large gap. The data for each channel are usually captured by different CCDs and are therefore reduced separately.

@@

Finally, I consider the handling of outliers in the data. These may be in the form of cosmic ray spikes, improper calibration of the data, telluric features, or simply poorly modelled spectral regions. Treatment of these artefacts is achieved using a Gaussian mixture model: a combination of two models. In the mixture model, the data are fit by the sum of amplitudes ($1 - P_o$ and $P_o$, respectively) of two distributions: the expected fluxes $E_i$, and a normal distribution centered along the continuum function $C_i$ with variance $s_{i}^2 + V_{o}$. This requires the inclusion of two (final) additional parameters: $P_o$ and $V_o$. The prior distribution function $p\left(V_o\right)$ requires $V_{o}$ to always be positive (Equation \ref{eq:default_priors}), and as such the outlier distribution will \textit{always} have a larger variance. Distributions of smaller variance are more informative, so conceptually a fit to the expected fluxes $E_{i}$ is generally preferred wherever possible. For brevity I define $\bm{\kappa} \equiv (\bm{\psi},\{z,\sigma_s,\{c_k\}_{k=0}^{m},f\}_{0}^{N_{c}})$, and the likelihood for the mixture model is given by

@@

\noindent{}where $p_{model}$ refers to $p$ in Equation \ref{eq:p_model} and

\begin{equation}
p_{outlier}\left(F_i|\lambda_i,\sigma_i,\bm{\kappa},V_{o},P_o\right) = \frac{1}{\sqrt{2\pi\left(s_{i}^2 + V_{o}^2\right)}} \exp\left(-\frac{[F_i - C_i]^2}{2\left[s_{i}^2 + V_{o}^2\right]}\right)
\end{equation}

@@

\begin{equation}
\mathcal{L} = \prod_{i=1}^{N} \left[ \frac{1-P_o}{\sqrt{2\pi{}s_{i}^2}}\,\exp\,\left(-\frac{[F_i - E_i]^2}{2s_{i}^{2}}\right) + \frac{P_o}{\sqrt{2\pi\left[s_{i}^2 + V_o\right]}}\,\exp\,\left(-\frac{[F_i - C_i]^2}{2\left[s_{i}^{2} + V_o\right]}\right)\right]
\label{eq:full_likelihood}
\end{equation}

I define the full parameter space with $\bm{\theta} \equiv \left(\bm{\psi},\{z,\sigma_s,\{c_{b_k}\}_{k=0}^{m},f\}_{b=0}^{N_{c}},V_o,P_o\right)$. From Bayes theorem the posterior probability distribution for $\bm{\theta}$ (up to a constant) is given by
\begin{eqnarray}
\mathcal{P} & \propto & likelihood \times prior \nonumber \\
p(\bm{\theta}|\{F_i\}_{i=1}^{N}) & \propto & p(\{F_i\}_{i=1}^{N}|\bm{\theta})\,\times\,p(\bm{\theta})
\label{eq:probability}
\end{eqnarray}

\noindent{}where $p(\bm{\theta}|\{F_i\}_{i=1}^{N})$ is the probability $\mathcal{P}$ of $\bm{\theta}$ given the data (and given the model: e.g., see Section \ref{sec:sun}), $p(\{F_i\}_{i=1}^{N}|\bm{\theta})$ is our previously defined likelihood function $\mathcal{L}$ (Equation \ref{eq:full_likelihood}), and $p(\bm{\theta})$ is the prior probability distribution. Priors are discussed in more detail in Section \ref{sec:mcmc}.

@@

With all of the combined effects there are 16 free parameters including $\bm{\psi}$, our astrophysical quantities of interest. For the examples presented in Section \ref{sec:examples}, this requires interpolation between 92 billion pixels in four dimensions. While the description might appear daunting, the problem is tractable, numerically efficient, and easy to configure. By default \sick{} numerically solves this problem in three sequential steps, which are described in the following sections: scattering, optimisation, and Monte-Carlo Markov Chain (MCMC) sampling.

@@

We seek to maximise $\mathcal{P}$ (or in practice, $\log\mathcal{P}$) and calculate the posterior probability distributions for $\bm{\theta}$ given the data. In the first step I calculate the log-probability $\log{(\mathcal{P})}$ for $N_{sample}$ randomly drawn points from all over the astrophysical parameter space $\bm{\psi}$. Since \sick{} allows for arbitrarily large parameter spaces in $N_{D}$ dimensions, initially sampling the parameter space $N_{sample}$ times provides a coarse landscape of probability. If the probability distribution function were smoothly distributed across all $\bm{\psi}$, or the optimisation step was sufficiently robust for all potential applications, then in principle \textit{any} single point would be an adequate starting guess. However there is no way of knowing \textit{a priori} how smooth the posterior distribution will be for a given problem.

The $N_{samples}$ points are uniformly drawn in $\bm{\psi}$ by default, unless priors are specified (see Section \ref{sec:mcmc}). A model spectrum $S(\bm{\psi})$ is interpolated for each point, which is used to estimate the redshift and continuum parameters. After smoothing the model fluxes $I_j$ by $\sigma_s$ (as drawn from an explicit prior, or estimated by $\left|\mathcal{N}\left(0, 1\right)\right|$) the smoothed spectrum is cross-correlated with the data to yield a redshift estimate. Continuum parameters $\{c_k\}_{k=0}^{m}$ are similarly estimated by fitting a function (a polynomial in this case) to the data divided by the smoothed flux at each $\lambda_i$. Estimating continuum and redshift provides us with a reasonable guess of the probability for any randomly scattered point $\bm{\psi_i}$. If outlier modelling is included in the model then $P_o$ is distributed as $\mathcal{U}\left(0, 1\right)$ by default, and $V_o$ is estimated by $\mathcal{N}\left(\widetilde{F_i}, \frac{1}{2}\widetilde{F_i}\right)$. This procedure is \textit{only} employed for the initial random scattering stage: it is \textit{not} applicable for the MCMC phase.

@@

After the random scattering step, the most probable $\bm{\theta}$ point is used as an initial guess for numerical optimisation. A number of suitable minimisation algorithms are available in \sick{} through the SciPy \citep{scipy} optimization module. The Nelder-Mead algorithm \citep{nelder-mead} is employed by default, and the full suite of optimisation algorithms available is listed in the online documentation. Unlike other optimisation techniques, the Nelder-Mead algorithm does not approximate first- or second-order derivatives, and can consequently be less efficient than other approaches. However it is robust in high-dimensional space, even in the presence of substantial noise, and has been successfully employed in a wide range of engineering and scientific problems. The reader is encouraged to experiment with other optimisation algorithms if the Nelder-Mead algorithm proves unsuitable or takes an untenable amount of time. Since these optimisation algorithms are minimisation techniques and I seek to maximise the log-probability $\log{\left(\mathcal{P}\right)}$, I numerically optimise the parameters $\bm{\theta}$ by minimising the negative log-probability $-\log{\left(\mathcal{P}\right)}$.

@@

The random scattering and numerical optimisation steps efficiently provide an accurate estimate of the optimal parameters $\bm{\theta}$. Once the optimisation step is complete, \sick{} employs the affine-invariant ensemble sampler proposed by \citet{goodman;weare}, and implemented by \citet{emcee}. The Metropolis-Hastings MCMC algorithm is employed by default. \sick{} allows for the model settings to be specified in a human-readable \textsc{yaml}- or \textsc{json}-formatted configuration file\footnote{The reader is referred to the online documentation for an example.}, where the number of Goodman \& Weare walkers can be specified, as well as the number of samples to perform. When the optimisation step is used, the initial points are taken from a small multi-dimensional ball around the optimised parameters $\bm{\theta_{opt}}$. If the optimisation step is not performed, initial values for the MCMC walkers are drawn in the same way as described for the scattering process.

Priors represent our initial knowledge about a particular parameter before looking at the data, and are necessary for any Bayesian analysis. A number of different prior distributions can be specified by the user in the \sick{} model configuration file. When no prior is explicitly specified, the following uninformative prior distributions are assumed (for all channels, where appropriate):

@@

A consequence of allowing irregular model grids is that occasionally a model spectrum cannot be interpolated for some values of $\bm{\psi}$, even if they fall within $\left(\min\left[\bm{\psi}\right], \max\left[\bm{\psi}\right]\right)$. In these cases $p\left(\bm{\psi}\right) = 0$ and thus $\log\left(\mathcal{P}\right) = -\infty$.
